---
title: "Signal and Noise Subgroups"
author: "Andrew Zimolzak"
date: "2022-10-09"
output: pdf_document
---

When a variable has no effect in most groups, but has an effect in one group,
it can look like no effect in the whole population.

```{r libraries, include=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
```

```{r functions, include=FALSE}
noisy_line <- function(slope, intercept, n, Subgroup) {
  
  n = n - 1
  
  data.frame(x = 0:n / n * 10, eps = rnorm(n + 1, 0, 1)) %>%
    mutate(y = slope * x + intercept + eps,
           Subgroup = Subgroup) -> df
  
  return(df)
  
}
```

```{r three-populations, include=FALSE}
nlow = 100
nmed = 100
nhi = 20

grp1 <- noisy_line(0, 3.5, nlow, '1 (Low risk)')
grp2 <- noisy_line(0, 6.5, nmed, '2 (Medium risk)')
grp3 <- noisy_line(1, 0, nhi, '3 (High risk)')

df <- rbind(grp1, grp2, grp3)

```

We will simulate a population made up of `r nlow` low-risk people, `r nmed`
medium-risk people, and `r nhi` high-risk people. The outcome $y$ has some random
variation. The data are constructed so that the exposure $x$ has no effect on $y$, *except* in the high-risk group, where it has a linear, positive effect.




# Effect of $x$ in the *whole* population

We will fit a simple model where $y = f(x)$, disregarding subgroups entirely.

```{r non-obvious-plot, echo=FALSE}

bad_lm <- lm(df$y ~ df$x)

ggplot(df, aes(x=x, y=y)) +
  geom_point() +
  geom_abline(slope = bad_lm$coefficients[2], intercept = bad_lm$coefficients[1])

summ = summary.lm(bad_lm)

```

Fitted the following linear model $\hat{y} = \beta_1 x + \beta_0$:

Slope = `r bad_lm$coefficients[2]`, $P=$ `r summ$coefficients[2,4]`

Intercept = `r bad_lm$coefficients[1]`, $P=$ `r summ$coefficients[1,4]`

$R^2=$ `r summ$r.squared` 

It appears that $x$ has little to no effect on $y$ (unimpressive $P$ for the slope
term, and unimpressive $R^2$).




# Effect in high-risk population only

We will fit a simple model but *only to one subgroup.*

```{r obvious-plot, echo=FALSE}
df %>%
  filter(Subgroup == '3 (High risk)') -> hr

my_lm <- lm(hr$y ~ hr$x)

summ = summary.lm(my_lm)

ggplot(df, aes(x=x, y=y)) +
  geom_point(aes(color=Subgroup)) +
  geom_abline(slope = my_lm$coefficients[2], intercept = my_lm$coefficients[1])

```

Fitted the following linear model $\hat{y} = \beta_1 x + \beta_0$:

Slope = `r my_lm$coefficients[2]`, $P=$ `r summ$coefficients[2,4]`

Intercept = `r my_lm$coefficients[1]`, $P=$ `r summ$coefficients[1,4]`

$R^2=$ `r summ$r.squared` 

If we separate the subgroups, we see that $x$ has a significant effect on $y$ in 
one particular subgroup (low $P$ for the slope term, and high $R^2$).




# Simple model adjusting for risk level

This will fit a multivariable (specifically two-variable) model to predict $y$ as a function of $x$ and subgroup both, *without* considering interaction between the two.

This table shows regression coefficient estimates and $P$ values:

```{r more-models, echo=FALSE}
mod2factor <- lm(data = df, y ~ x + Subgroup)
modinter <- lm(data = df, y ~ x + Subgroup + (x * Subgroup))

sm2 <- summary.lm(mod2factor)
smi <- summary.lm(modinter)

sm2$coefficients[, c(1,4)] %>% as.table() %>% kable()
```

$R^2 =$ `r sm2$r.squared`

## Interpretation

Intercept:
: What is the best guess for $y$ (for the reference subgroup)?

x:
: What should we add to that, for each 1.0 increase in $x$? (Often not significant)

Subgroup 2:
: What should we add for points in subgroup 2?

Subgroup 3:
: What should we add for points in subgroup 3, the high-risk group? (Often significant for my default random data, because mean of group 1 $\approx 3.5$, mean of group 1 $\approx 6.5$, and mean of group 3 $\approx 5$.)

*Note:* The parameter for $x$ (the slope) is estimated the same for all subgroups.




# Model including exposure:risk interaction

Finally, we fit a multivariable of $y$ as a function of $x$ and subgroup both, *plus* an interaction term. This should do approximately the same thing as fitting three separate one-variable models, one per each stratum, but now we are combining these into an "all-in-one" model.

```{r echo=FALSE}
# smi$coefficients[, c(1,4)]  %>% as.data.frame() -> outdf
# hux(outdf) %>%
#   add_rownames()

smi$coefficients[, c(1,4)] %>% as.table() %>% kable()
```

$R^2 =$ `r smi$r.squared`

## Interpretation

First, is the $R^2$ better in the simple model or the interaction model?

Intercept:
: What is the best guess for $y$ (for the reference subgroup)?

x:
: What should we add to that, for each 1.0 increase in $x$ (In other words, the slope for the reference subgroup)?

Subgroup 2:
: What should we add (to intercept) for points in subgroup 2?

Subgroup 3:
: What should we add (to intercept) for points in subgroup 3, the high-risk group? (Interpret in context of the slope, below.)

x:Subgroup 2 interaction:
: Effect of changing group from 1 to 2 on the slope. (If significant, then group 2 has a different slope from group 1)

x:Subgroup 3 interaction:
: Effect of changing group from 1 to 3 on the slope. (If significant, then group 3 has a different slope from group 1)

Note that this final $R^2$ is for the "whole" model, accounting for all 3 subgroups, so if it is not as "impressive" as the one-subgroup model $R^2$ above, it may be "diluted" by the other two subgroups.
