---
title: "Signal and Noise Subgroups"
author: "Andrew Zimolzak"
date: "2022-10-09"
output: pdf_document
---

When a variable has no effect in most groups, but has an effect in one group,
it can look like no effect in the whole population.

```{r libraries, include=FALSE}
library(ggplot2)
library(dplyr)
```

```{r functions, include=FALSE}
noisy_line <- function(slope, intercept, n, Subgroup) {
  
  n = n - 1
  
  data.frame(x = 0:n / n * 10, eps = rnorm(n + 1, 0, 1)) %>%
    mutate(y = slope * x + intercept + eps,
           Subgroup = Subgroup) -> df
  
  return(df)
  
}
```

```{r three-populations, include=FALSE}
nlow = 100
nmed = 100
nhi = 20

grp1 <- noisy_line(0, 3.5, nlow, 'Low risk')
grp2 <- noisy_line(0, 6.5, nmed, 'Medium risk')
grp3 <- noisy_line(1, 0, nhi, 'High risk')

df <- rbind(grp1, grp2, grp3)

```

We will simulate a population made up of `r nlow` low-risk people, `r nmed`
medium-risk people, and `r nhi` high-risk people. The outcome $y$ has some random
variation. The exposure $x$ has no effect on $y$, *except* in the high-risk group,
where it has a linear, positive effect.




# Effect in whole population

```{r non-obvious-plot, echo=FALSE}

bad_lm <- lm(df$y ~ df$x)

ggplot(df, aes(x=x, y=y)) +
  geom_point() +
  geom_abline(slope = bad_lm$coefficients[2], intercept = bad_lm$coefficients[1])

summ = summary.lm(bad_lm)

```

Fitted the following linear model $\hat{y} = \beta_1 x + \beta_0$:

Slope = `r bad_lm$coefficients[2]`, $P=$ `r summ$coefficients[2,4]`

Intercept = `r bad_lm$coefficients[1]`, $P=$ `r summ$coefficients[1,4]`

$R^2=$ `r summ$r.squared` 

It appears that $x$ has little to no effect on $y$ (unimpressive $P$ for the slope
term, and unimpressive $R^2$).




# Effect in high-risk population only

```{r obvious-plot, echo=FALSE}
df %>%
  filter(Subgroup == 'High risk') -> hr

my_lm <- lm(hr$y ~ hr$x)

summ = summary.lm(my_lm)

ggplot(df, aes(x=x, y=y)) +
  geom_point(aes(color=Subgroup)) +
  geom_abline(slope = my_lm$coefficients[2], intercept = my_lm$coefficients[1])

```

Fitted the following linear model $\hat{y} = \beta_1 x + \beta_0$:

Slope = `r my_lm$coefficients[2]`, $P=$ `r summ$coefficients[2,4]`

Intercept = `r my_lm$coefficients[1]`, $P=$ `r summ$coefficients[1,4]`

$R^2=$ `r summ$r.squared` 

If we separate the subgroups, we see that $x$ has a significant effect on $y$ in 
one particular subgroup (low $P$ for the slope term, and high $R^2$).
