---
title: "Signal and Noise Subgroups"
author: "Andrew Zimolzak"
date: "2022-10-09"
output: html_document
runtime: shiny
---

When a variable has no effect in most groups, but has an effect in one group,
it can look like no effect in the whole population.

```{r libraries, include=FALSE}
library(ggplot2)
library(dplyr)
library(shiny)
```

```{r functions, include=FALSE}
noisy_line <- function(slope, intercept, n, Subgroup, noise=1) {
  
  n = n - 1
  
  data.frame(x = 0:n / n * 10, eps = rnorm(n + 1, 0, noise)) %>%
    mutate(y = slope * x + intercept + eps,
           Subgroup = Subgroup) -> df
  
  return(df)
  
}
```




## Settings

Choose how you want your populations to look.

```{r inputs, echo=FALSE}
inputPanel(
  # selectInput("do_subgroup", label = "Analyze high-risk only:",
  #             choices = c(TRUE, FALSE), selected = FALSE),
  sliderInput("nlow", label = "N of low-risk:",
              min = 50, max = 200, value = 100, step = 1),
  sliderInput("nmed", label = "N of medium-risk:",
              min = 50, max = 200, value = 100, step = 1),
  sliderInput("nhi", label = "N of high-risk:",
              min = 10, max = 40, value = 20, step = 1)
)

inputPanel(
  sliderInput("mlow", label = "Slope low-risk:",
              min = -0.1, max = 0.1, value = 0, step = 0.01),
  sliderInput("mmed", label = "Slope medium-risk:",
              min = -0.1, max = 0.1, value = 0, step = 0.01),
  sliderInput("mhi", label = "Slope high-risk:",
              min = 0.1, max = 2, value = 1, step = 0.1)
)

inputPanel(
  sliderInput("blow", label = "Intercept low-risk:",
              min = 2.5, max = 5, value = 3.5, step = 0.1),
  sliderInput("bmed", label = "Intercept medium-risk:",
              min = 5, max = 10, value = 7.5, step = 0.1),
  sliderInput("bhi", label = "Intercept high-risk:",
              min = 0, max = 5, value = 0, step = 0.1)
)

inputPanel(
  sliderInput("noise", label = "Noise:",
              min = 0, max = 5, value = 1, step = 0.1)
)
```


```{r function-plot, echo=FALSE}
make_rand_plot <- function(split_subgroups = TRUE, return_what) {
  set.seed(42)
  grp1 <- noisy_line(input$mlow, input$blow, input$nlow, '1 (Low risk)', input$noise)
  grp2 <- noisy_line(input$mmed, input$bmed, input$nmed, '2 (Medium risk)', input$noise)
  grp3 <- noisy_line(input$mhi, input$bhi, input$nhi, '3 (High risk)', input$noise)
  
  df <- rbind(grp1, grp2, grp3)
  
  if(split_subgroups) {
    df %>% filter(Subgroup == '3 (High risk)') -> subg
    model <- lm(subg$y ~ subg$x)
    myplot <- ggplot(df, aes(x=x, y=y)) +
      geom_point(aes(color = Subgroup)) +
      geom_abline(slope = model$coefficients[2], intercept = model$coefficients[1])
  }
  else {
    model <- lm(df$y ~ df$x)
    myplot <- ggplot(df, aes(x=x, y=y)) +
      geom_point() +
      geom_abline(slope = model$coefficients[2], intercept = model$coefficients[1])
  }
  
  summ = summary.lm(model)
  
  # multivariable
  mod2factor <- lm(data = df, y ~ x + Subgroup)
  modinter <- lm(data = df, y ~ x + Subgroup + (x * Subgroup))

  sm2 <- summary.lm(mod2factor)
  smi <- summary.lm(modinter)

  sm2$coefficients[, c(1,4)] %>% as.table() -> t2
  smi$coefficients[, c(1,4)] %>% as.table() -> ti

  if(return_what == 'plot'){
    return(myplot)
  }
  else if(return_what == 'm'){
    return(model$coefficients[2])
  }
  else if(return_what == 'b'){
    return(model$coefficients[1])
  }
  else if(return_what == 'pm'){
    return(summ$coefficients[2,4])
  }
  else if(return_what == 'pb'){
    return(summ$coefficients[1,4])
  }
  else if(return_what == 'r'){
    return(summ$r.squared)
  }
  else if(return_what == 't2'){
    return(t2)
  }
  else if(return_what == 'ti'){
    return(ti)
  }
  else if(return_what == 'r2'){
    return(sm2$r.squared)
  }
  else if(return_what == 'ri'){
    return(smi$r.squared)
  }
}
```




## All subgroups mixed

```{r final, echo=FALSE}
renderPlot({
  make_rand_plot(split_subgroups = FALSE, return_what = 'plot')
})
```

Fitted the following linear model $\hat{y} = \beta_1 x + \beta_0$:

Slope = `r reactive({make_rand_plot(split_subgroups = FALSE, return_what = 'm')})`,
$P=$ `r reactive({make_rand_plot(split_subgroups = FALSE, return_what = 'pm')})`

Intercept = `r reactive({make_rand_plot(split_subgroups = FALSE, return_what = 'b')})`,
$P=$ `r reactive({make_rand_plot(split_subgroups = FALSE, return_what = 'pb')})`

$R^2$ = `r reactive({make_rand_plot(split_subgroups = FALSE, return_what = 'r')})`




## Only one subgroup

```{r final2, echo=FALSE}
renderPlot({
  make_rand_plot(split_subgroups = TRUE, return_what='plot')
})
```

Fitted the following linear model $\hat{y} = \beta_1 x + \beta_0$:

Slope = `r reactive({make_rand_plot(split_subgroups = TRUE, return_what = 'm')})`,
$P=$ `r reactive({make_rand_plot(split_subgroups = TRUE, return_what = 'pm')})`

Intercept = `r reactive({make_rand_plot(split_subgroups = TRUE, return_what = 'b')})`,
$P=$ `r reactive({make_rand_plot(split_subgroups = TRUE, return_what = 'pb')})`

$R^2$ = `r reactive({make_rand_plot(split_subgroups = TRUE, return_what = 'r')})`




## Simple model adjusting for risk level

This will fit a multivariable (specifically two-variable) model to predict $y$ as a function of $x$ and subgroup both, *without* considering interaction between the two.

This table shows regression coefficient estimates and $P$ values:


```{r echo=FALSE}
reactive({
  make_rand_plot(return_what = 't2')
})
```

$R^2 =$ `r reactive({make_rand_plot(return_what = 'r2')})`

### Interpretation

Intercept:
: What is the best guess for $y$ (for the reference subgroup)?

x:
: What should we add to that, for each 1.0 increase in $x$? (Often not significant)

Subgroup 2:
: What should we add for points in subgroup 2?

Subgroup 3:
: What should we add for points in subgroup 3, the high-risk group? (Often significant for my default random data, because mean of group 1 $\approx 3.5$, mean of group 1 $\approx 6.5$, and mean of group 3 $\approx 5$.)

*Note:* The parameter for $x$ (the slope) is estimated the same for all subgroups.




## Model including exposure:risk interaction

Finally, we fit a multivariable of $y$ as a function of $x$ and subgroup both, *plus* an interaction term. This should do approximately the same thing as fitting three separate one-variable models, one per each stratum, but now we are combining these into an "all-in-one" model.

```{r echo=FALSE}
reactive({
  make_rand_plot(return_what = 'ti')
})

```

$R^2 =$ `r reactive({make_rand_plot(return_what = 'ri')})`

### Interpretation

First, is the $R^2$ better in the simple model or the interaction model?

Intercept:
: What is the best guess for $y$ (for the reference subgroup)?

x:
: What shouldwe add to that, for each 1.0 increase in $x$ (In other words, the slope for the reference subgroup)?

Subgroup 2:
: What should we add (to intercept) for points in subgroup 2?

Subgroup 3:
: What should we add (to intercept) for points in subgroup 3, the high-risk group? (Interpret in context of the slope, below.)

x:Subgroup 2 interaction:
: Effect of changing group from 1 to 2 on the slope. (If significant, then group 2 has a different slope from group 1)

x:Subgroup 3 interaction:
: Effect of changing group from 1 to 3 on the slope. (If significant, then group 3 has a different slope from group 1)

Note that this final $R^2$ is for the "whole" model, accounting for all 3 subgroups, so if it is not as "impressive" as the one-subgroup model $R^2$ above, it may be "diluted" by the other two subgroups.
